Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-04 examples.tex
+

 In case of absence, \textsf{pte\_initialize} is called for a physical-zeroed page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiom as it is not part of \textsf{pte} (virtual-memory-management) module) to be backing the physical frame address of the entry -- Specification Line 35 in Figure \ref{fig:calltopteinitialize}. Since we are using \textsf{pte\_initialize} for page-table address allocation, it also ensures us that newly allocated physicall address is in the identity mapping map $\Xi$ -- Line X in Figure \ref{fig:calltopteinitialize}.
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{\_} \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_}\ast\sumapaces\rtv\delta }_{\rtv}$
;;void pte_initialize(pte_t **table) {
;;set up the stack    
;;allocate a full zeroed page for 512 8-byte entries 
;;pte_t *local = kalloc();
callq  81 <pte_initialize+0x11>	7d: R_X86_64_PLT32	kalloc-0x4
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = None \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_} \ast \sumapaces\rtv\delta }_{\rtv}$
$\sqsubseteq$
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\cup\{\mathsf{nextpaddr}\},m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = Z \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_} \ast \sumapaces\rtv\delta }_{\rtv}$  
mov    rax,-0x10[rbp] 
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\cup\{\mathsf{nextpaddr}\},m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = Z \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_}\ast \sumapaces\rtv\delta }_{\rtv}$  
;;*table = local;
mov    -0x10[rbp],rax
mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\cup\{\mathsf{nextpaddr}\},m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = Z \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\mathsf{nextpaddr}} \ast \sumapaces\rtv\delta }_{\rtv}$  
;;return;
;;..
;;clean up the stack
\caption{Allocating a physical page }
\label{pteinitializespec}
\mytodo{give line numbers andrelated post here and pteinitcal }
 There two important aspects on the specification after the call to \textsf{pte\_initialize} -- Specification in Figure \ref{fig:pteinitializespec}. First, one might have expected to see a physical pointsto for the fresly page-table address  ($\mathsf{nextpaddr} \mapsto_{\mathsf{p}} \mathsf{w64\_0}$) after the call to \textsf{pte\_initialize} but we do not have this physical pointsto resource. Second, we see the identity mapping for \textsf{nextpaddr} -- $\ulcorner \Xi !! \textsf{nextpaddr} = \textsf{ Some Z} \urcorner$ -- mapped to a different value (\textsf{Z}) than the identity mapping for the physical address \textsf{entry} is (\textsf{V}). The first aspect is important for not invalidating the soundness: \textsf{pte\_initialize} extends the identity map with the freshly allocated \textsf{nextpaddr}, and must not reveal the ownership of the physical pointsto in the post-condition (Line X in Figure \ref{}) because it would create a duplication of the full-ownership of the physical resource in the post-condition would potentially coexists with the one that can be obtained from the side-condition (*) inside the extended invariant in Figure \ref{fig:peraspaceinvariant_with_p2v_extension} with the capability token $\ulcorner \Xi !! \textsf{nextpaddr} = \textsf{ Some Z} \urcorner$. Although, the first aspect recovers the soundness while preserving the knowledge of existence of the identity mapping for the fresly allocated physical page-table address, we still need the knowledge that the fresly allocated address references to a \textit{zeroed} page that an explicitly revealed physical pointsto would give -- $\mathsf{nextpaddr} \mapsto_{\mathsf{p}} \mathsf{w64\_0}$. Since a value of an identity mapping inside the invariant is existentially quantified (\textsf{val} in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}), we would loose the knowledge of identity mapping with zeroed value. To carry this knowledge, we distinguish a random value of an identity mapping -- capability representing the existence of the random value (\textsf{V}) inside the invariant for the address \textsf{entry} at Line X in Figure \ref{} -- from the capability representing the zeroed identity-mapping value zeroed (\textsf{Z}). As we do for the address \textsf{nextpaddr}, when we use the capability with value \textsf{Z} for the identity-map lookup ($\Xi !! \textsf{nextpaddr} = \textsf{ Some Z}$), we know that the value existentially quantified inside the invariant is zeroed -- \textsf{val == w64\_0}.
  \subsubsection{Kernels Utilizing \textsf{P2V}}
  \label{sec:p2v}
 Once we make sure that the frame of the page-table entry ($\textsf{entry}\mapsto\textsf{pfn}$)is backed with an allocated physical page, we must obtain the virtual address of the next level-page table (Line 46 in \ref{calltopteinitialize}). To do so, as for obtaining any virtual page-table address, we, again, need to open the invariant look up the \textit{identity mapping} for the physical address ($\textsf{entry}\mapsto \textsf{pfn}$).

>>>>

 In case of absence, \textsf{pte\_initialize} is called (Line 31 in Figure \ref{fig:calltopteinitialize}) for a physical-zeroed page (utilizing the page-allocator's \textsf{kalloc} -- currently the only axiomatized call in the the proof of \textsf{pte\_initialize} (Line 9 in Figure \ref{pteinitializespec}, it is not part of \textsf{pte} (virtual-memory-management) module). 
\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{\_} \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_}\ast\sumapaces\rtv\delta }_{\rtv}$
;;void pte_initialize(pte_t **table) {
...
;;set up the stack    
;;allocate a full zeroed page for 512 8-byte entries 
;;pte_t *local = kalloc();
callq  81 <pte_initialize+0x11>	7d: R_X86_64_PLT32	kalloc-0x4
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = None \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_} \ast \sumapaces\rtv\delta }_{\rtv}$
$\sqsubseteq$
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\cup\{\mathsf{nextpaddr}\},m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry\_val}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = Z \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_} \ast \sumapaces\rtv\delta }_{\rtv}$  
mov    rax,-0x10[rbp] 
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\cup\{\mathsf{nextpaddr}\},m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = Z \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_}\ast \sumapaces\rtv\delta }_{\rtv}$  
;;*table = local;
mov    -0x10[rbp],rax
mov    -0x8[rbp],rdi
mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\cup\{\mathsf{nextpaddr}\},m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi !! \mathsf{nextpaddr} = Z \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\mathsf{nextpaddr}} \ast \sumapaces\rtv\delta }_{\rtv}$  
;;return;
;;..
;;clean up the stack
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\end{figure}

Since we are using \textsf{pte\_initialize} for page-table address allocation, we must relate this newly allocated physical address to the identity mapping map $\Xi$ -- Specification Lines 33 and 35 in Figure \ref{fig:calltopteinitialize}. There are two important aspects on relating a freshly allocated page-table physical address to the identity-mapping which we focus in more detail separately in the specification of \textsf{pte\_initialize} in Figure \ref{fig:pteinitializespec}. First, one might have expected to see a physical pointsto for the fresly page-table address  ($\mathsf{nextpaddr} \mapsto_{\mathsf{p}} \mathsf{w64\_0}$) after the call to \textsf{pte\_initialize} but we do not have this physical pointsto resource. Second, we see the identity mapping for \textsf{nextpaddr}, first not appearing in the identity map (Specification Line 11 in Figure \ref{fig:pteinitializespec}), then as inserted into the identity-map with a value \textsf{Z} ($\Xi \cup \{\mathsf{nextpaddr}\}$) together with the fragmental capability  -- $\ulcorner \Xi \; !!\; \textsf{nextpaddr} = \textsf{ Some Z} \urcorner$ -- at Lines 13 and 14 in Figure \ref{fig:pteinitializespec} . The value type \textsf{Z}, unlike the value type \textsf{V} denoting the random value in Line 8 in Figure \ref{fig:calltopteinitialize}, denotes \textsf{zeroed} value.

The first aspect is important for not invalidating the soundness: \textsf{pte\_initialize} extends the identity map with the freshly allocated \textsf{nextpaddr}, and must not reveal the ownership of the physical pointsto in the post-condition (Line 33 and 35 in Figure \ref{fig:calltopteinitialize}, Lines 22-23 in Figure \ref{fig:pteinitializespec}) because it would create unsound coexistence of the full-ownership of the physical pointsto resource (a hypothetical $\mathsf{nextpaddr} \mapsto_{\mathsf{p,1}} w0_64$) and fragmental (potentially full based on the side-condition *) physical pointsto resource that can be obtained from the identity map inside the extended invariant in Figure \ref{fig:peraspaceinvariant_with_p2v_extension} with the capability token $\ulcorner \Xi !! \textsf{nextpaddr} = \textsf{ Some Z} \urcorner$. Although, the first aspect -- not explicitly revealing the physical pointsto resource for the freshly allocated page-table address but providing access through the identity map inside the invariant with access-capability tokens -- recovers the soundness argument expalined, we still need the knowledge that the fresly allocated address references to a \textit{zeroed} page. The invariant preserves this information through the constraint on relating the type (\textsf{T}) of mapped data -- \textsf{T = Z} for zeroed or \textsf{T = V} for random page value -- to the value of existentially quantified value -- \textsf{val} -- in Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. So, for accessing the physical pointsto resource to the zeroed value referenced by \textsf{nextpaddr}, we need to have the fragmental ownership of the identity-map $\Xi !! \textsf{nextpaddr} = \textsf{ Some Z}$ which ensures and carries the \textit{being-zeroed} knowledge when we open and access the the existentially quantified value inside the invariant.
  \subsubsection{Kernels Utilizing \textsf{P2V}}
  \label{sec:p2v}
 Once we make sure that the frame of the page-table entry ($\textsf{entry}\mapsto\textsf{pfn}$)is backed with an allocated physical page, we must obtain the virtual address of the next level-page table (Line 46 in \ref{calltopteinitialize}). To do so, as for obtaining any virtual page-table address, we, again, need to open the invariant look up the \textit{identity mapping} for the physical address ($\textsf{entry}\mapsto \textsf{pfn}$).
